Concurrent Programming with CXL

### 2020-06-23 book/paper.tex
Adjusting to new formatting

\begin{itemize}
\item Chapter~\ref{ch:cxlintro} introduces the CXL programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item Chapter~\ref{ch:concurrent} illustrates the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter.
\item Chapter~\ref{ch:cxlmachine} presents the
CXL virtual machine to understand the problem
underlying concurrency better.
\item Chapter~\ref{ch:critical} introduces the concept of a
\emph{critical section} and presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
elegant solution to implementating a critical section.
\item Chapter~\ref{ch:method} gives some more details on the CXL
language.
\item Chapter~\ref{ch:spinlock} introduces ``hardware'' locks
for implemented critical sections.
\item Chapter~\ref{ch:synch} presents CXL modules
that supports locks and other synchronization primitives.
\item Chapters~\ref{ch:rdwrbusy} and~\ref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
Chapter~\ref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item Chapter~\ref{ch:deadlock} describes \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item Chapter~\ref{ch:semaphore} introduces \emph{semaphores},
a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the
\emph{bounded buffer problem} (aka \emph{producer/consumer problem})
and various solutions.
\item Chapter~\ref{ch:starvation} talks about \emph{starvation}:
the problem that in some
synchronization approaches processes may not be able to get access to a
resource they need.
\item Chapter~\ref{ch:monitors} presents
\emph{monitors} and \emph{condition variables},
another approach to process synchronication.
\item Chapter~\ref{ch:deadlockprevention} revisits deadlocks and how to
avoid them.
\item Chapter~\ref{ch:actor} presents the \emph{actor model} as an approach to
synchronization.
\item Chapter~\ref{ch:nonblocking} introduces \emph{non-blocking} or
\emph{wait-free} synchronization algorithms,
which prevent processes waiting for one another more than a bounded number of
steps.
\item Chapter~\ref{ch:barrier} describes \emph{barrier synchronization},
useful in high-performance computing applications such as parallel simulations.
\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\end{itemize}

@@

\begin{figure}
\begin{center}
\includegraphics[width=4in]{figures/mutex-crop.pdf}
\end{center}
\caption{High-level state diagram specification of mutual exclusion with up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes in the critical section.}
\label{fig:mutex}
\end{figure}

@@

\begin{figure}
\begin{center}
\includegraphics[width=2in]{figures/rdwr-crop.pdf}
\end{center}
\caption{High-level state diagram specification of reader/writer locks with
up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes reading in the critical section; the third is the number of
processes writing in the critical section.}
\label{fig:rdwr}
\end{figure}

@@

Figure~\ref{fig:rwbusy} presents a solution that uses a
single (ordinary) lock and two counters: one that maintains the number
of readers and one that maintains the number of writers.
We will call the lock the \emph{mutex} to distinguish it clearly from
the reader/writer lock that we are implementing.
The mutex is used to protect shared access to the counters.
The program shows a process that executes in a loop.
Each time, it decides whether to read or write.
The critical section is spread between two labels:
readers access \texttt{@rcs} and writers access \texttt{@wcs}.
The specification is that if a reader is at label \texttt{@rcs},
no writer is allowed to be at label \texttt{@wcs}.  Vice versa, if
a writer is at label \texttt{@wcs}, no reader is allowed to be at
label \texttt{@rcs} \emph{and} there cannot be another writer at
label \texttt{@wcs}.  Figure~\ref{fig:rdwr} shows a high-level
specification for two processes.

@@

\glsadd{busy waiting}
While correct, it is not considered a good solution.
The solution is an example of what is called \emph{busy-waiting}
\index{busy waiting}
(aka \emph{spin-waiting}):
\index{spin waiting}
processes spin in a loop until some desirable application-level condition is met.
Figure~\ref{fig:inc} also has an example of busy-waiting: the \texttt{main}
process waits for the other two processes to finish by checking their \texttt{done}
flags.
The astute reader might wonder if obtaining the mutex itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the mutex is available
(see Figure~\ref{fig:spinlocks}).

However, there is a big difference.  As we pointed out, the processes that
are waiting for a spinlock are \emph{blocked}.
In most operating systems and programming language runtimes,
when a process acquires a mutex, the process is placed on a scheduling
queue and stops using CPU cycles until the mutex becomes available.
CXL supports this with the \texttt{synchS} module.
Busy waiting disables all this: even when using the \texttt{synchS}
module, readers and writers only get suspended temporarily in case
there is contention for the mutex.
A process trying to obtain a read or write lock in
Figure~\ref{fig:rwbusy} has to obtain the mutex repeatedly
until the read or write lock is available.

@@

\begin{quote}
\begin{itemize}
\item Do you see what the problem is?
\item Does it depend on \texttt{N}, the number of philosophers?
\end{itemize}
\end{quote}

We will discuss ways of preventing deadlock later (Chapter~\ref{ch:deadlockprevention}),
but one important
approach to preventing deadlock is to prevent a cycle from forming, with each
process waiting for the next process on the cycle.
We can do this by establishing an ordering among the
resources (in this case the forks) and, when needing more than one
resource, always acquiring them in order.  In the case of the philosopers,
they could prevent deadlock by always picking up the lower numbered fork
before the higher numbered fork, like so:

\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
    if left < right:
        lock(&forks[left]);
        lock(&(orks[right]);
    else:
        unlock(&forks[right]);
        unlock(&forks[left]);
    ;
\end{Verbatim}
\end{code}

\section*{Exercises}
\begin{problems}
\item Does it matter in what order the philosophers lay down their forks?
\item Figure~\ref{fig:bank} shows an implementation of a bank with various
accounts and transfers between those accounts.
Unfortunately, running the test reveals that it sometimes leaves unterminated
processes.  Can you fix the problem?
\end{problems}

@@

\chapter{Deadlock Prevention}
\label{ch:deadlockprevention}
\index{deadlock prevention}

@@

As introduced in Chapter~\ref{ch:deadlock}, deadlock is the situation when there are
a collection of processes waiting for one another to release one or more resources.
There are four conditions that must hold for deadlock to occur~\cite{CES71}:
\begin{enumerate}
\item \emph{Mutual Exclusion}: each resource can only be used by one process at a time:
\item \emph{Hold and Wait}: each process holds resources it already allocated while it
waits for other resources that it needs;
\item \emph{No Preemption}: Resources cannot be forcibly taken back from processes that
allocated them;
\item \emph{Circular Wait}: There exists a directed circular chain of processes, each waiting
to allocate a resource held by the next.
\end{enumerate}

@@

The \texttt{Preemption} approach suggested by Havender is to allow processes to back out.
While this could be done, this invariably leads to a busy waiting solution
where a process keeps obtaining locks and releasing them again until it
finally is able to get all of them.
The \emph{No Circular Waiting} approach to order the resources was already presented in
Chapter~\ref{ch:deadlock}

There is, however, another approach. which is sometimes called
\emph{deadlock avoidance}.
\index{deadlock avoidance}.
In the case of the Dining Philosophers, we want to avoid the situation where each
diner picks up a fork.  So if we can prevent more than four diners from starting to
eat at the same time, then we can avoid the conditions for deadlock from ever
happening.
Figure~\ref{fig:dinersavoid} demonstrates this concept.  It uses a semaphore to
restrict the number of diners at any time to four.

@@

>>>>

\begin{itemize}
\item Chapter~\ref{ch:cxlintro} introduces the CXL programming
language, as it provides the language for presenting synchronization
problems and solutions.
\item Chapter~\ref{ch:concurrent} illustrates the problem of
concurrent programming through a simple example in which two processes
are concurrently incrementing a counter.
\item Chapter~\ref{ch:cxlmachine} presents the
CXL virtual machine to understand the problem
underlying concurrency better.
\item Chapter~\ref{ch:critical} introduces the concept of a
\emph{critical section} and presents various flawed implementations
of critical sections to demonstrate that implementing a critical section
is not trivial.
\item Chapter~\ref{ch:peterson} introduces Peterson's Algorithm, an
elegant solution to implementating a critical section.
\item Chapter~\ref{ch:method} gives some more details on the CXL
language.
\item Chapter~\ref{ch:spinlock} introduces ``hardware'' locks
for implemented critical sections.
\item Chapter~\ref{ch:synch} presents CXL modules
that supports locks and other synchronization primitives.
\item Chapters~\ref{ch:rdwrbusy} and~\ref{ch:rdwr2} present the
abstraction and implementation of reader/writer
locks, which allow multiple processes in a critical section if they are
all only executing read-only accesses.
Chapter~\ref{ch:rdwrbusy} also talks about busy-waiting, which is
an undesirable approach to synchronize processes.
\item Chapter~\ref{ch:semaphore} introduces \emph{semaphores},
a generalization of locks
that is good not only for mutual exclusion but also for waiting for
certain application-level conditions.
\item Chapters~\ref{ch:bb} and~\ref{ch:sbs} introduce the
\emph{bounded buffer problem} (aka \emph{producer/consumer problem})
and various solutions.
\item Chapter~\ref{ch:starvation} talks about \emph{starvation}:
the problem that in some
synchronization approaches processes may not be able to get access to a
resource they need.
\item Chapter~\ref{ch:monitors} presents
\emph{monitors} and \emph{condition variables},
another approach to process synchronication.
\item Chapter~\ref{ch:deadlock} describes \emph{deadlock}
where a set of processes are indefinitely waiting for one another to
release a resource.
\item Chapter~\ref{ch:actor} presents the \emph{actor model} as an approach to
synchronization.
\item Chapter~\ref{ch:nonblocking} introduces \emph{non-blocking} or
\emph{wait-free} synchronization algorithms,
which prevent processes waiting for one another more than a bounded number of
steps.
\item Chapter~\ref{ch:barrier} describes \emph{barrier synchronization},
useful in high-performance computing applications such as parallel simulations.
\item Chapter~\ref{ch:abp} presents a problem and a solution to the distributed
systems problem of having two processes communicate reliably over an unreliable
network.
\end{itemize}

@@

\begin{figure}
\begin{center}
\includegraphics[width=4.5in]{figures/mutex-crop.pdf}
\end{center}
\caption{High-level state diagram specification of mutual exclusion with up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes in the critical section.}
\label{fig:mutex}
\end{figure}

@@

\begin{figure}
\begin{code}
\VerbatimInput[xleftmargin=5mm,numbers=left]{code/RWbusytest.cxl}
\end{code}
\caption{[RWbusytest.cxl] Test code for Figure~\ref{fig:rwbusy}.}
\label{fig:rwbusytest}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[width=2.5in]{figures/rdwr-crop.pdf}
\end{center}
\caption{High-level state diagram specification of reader/writer locks with
up to two processes.
The first number in a state gives the number of processes; the second number is the
number of processes reading in the critical section; the third is the number of
processes writing in the critical section.}
\label{fig:rdwr}
\end{figure}

@@

Figure~\ref{fig:rwbusy} presents a solution that uses a
single (ordinary) lock and two counters: one that maintains the number
of readers and one that maintains the number of writers.
Figure~\ref{fig:rwbusytest} shows how the code may be used.
We will call the lock the \emph{mutex} to distinguish it clearly from
the reader/writer lock that we are implementing.
The mutex is used to protect shared access to the counters.
The program shows a process that executes in a loop.
Each time, it decides whether to read or write.
The critical section is spread between two labels:
readers access \texttt{@rcs} and writers access \texttt{@wcs}.
The specification is that if a reader is at label \texttt{@rcs},
no writer is allowed to be at label \texttt{@wcs}.  Vice versa, if
a writer is at label \texttt{@wcs}, no reader is allowed to be at
label \texttt{@rcs} \emph{and} there cannot be another writer at
label \texttt{@wcs}.  Figure~\ref{fig:rdwr} shows a high-level
specification for two processes.

@@

\glsadd{busy waiting}
While correct, it is not considered a good solution.
The solution is an example of what is called \emph{busy-waiting}
\index{busy waiting}
(aka \emph{spin-waiting}):
\index{spin waiting}
processes spin in a loop until some desirable application-level condition is met.
Figure~\ref{fig:inc} also has an example of busy-waiting: the \texttt{main}
process waits for the other two processes to finish by checking their \texttt{done}
flags.

The astute reader might wonder if obtaining the mutex itself is an
example of busy-waiting.  After all, the CXL \texttt{synch} implementation
of \texttt{lock()} spins in a loop until the mutex is available
(see Figure~\ref{fig:spinlocks}).
However, there is a big difference.  As we pointed out, the processes that
are waiting for a spinlock are \emph{blocked}.
In most operating systems and programming language runtimes,
when a process acquires a mutex, the process is placed on a scheduling
queue and stops using CPU cycles until the mutex becomes available.
CXL supports this with the \texttt{synchS} module.
Busy waiting disables all this: even when using the \texttt{synchS}
module, readers and writers only get suspended temporarily in case
there is contention for the mutex.
A process trying to obtain a read or write lock in
Figure~\ref{fig:rwbusy} has to obtain the mutex repeatedly
until the read or write lock is available.

@@

@@

\begin{quote}
\begin{itemize}
\item Do you see what the problem is?
\item Does it depend on \texttt{N}, the number of philosophers?
\item Does it matter in what order the philosophers lay down their forks?
\end{itemize}
\end{quote}

% \chapter{Deadlock Prevention}
% \label{ch:deadlockprevention}
% \index{deadlock prevention}

@@

\noindent
There are four conditions that must hold for deadlock to occur~\cite{CES71}:
\begin{enumerate}
\item \emph{Mutual Exclusion}: each resource can only be used by one process at a time:
\item \emph{Hold and Wait}: each process holds resources it already allocated while it
waits for other resources that it needs;
\item \emph{No Preemption}: Resources cannot be forcibly taken back from processes that
allocated them;
\item \emph{Circular Wait}: There exists a directed circular chain of processes, each waiting
to allocate a resource held by the next.
\end{enumerate}

@@

The \emph{Preemption} approach suggested by Havender is to allow processes to back out.
While this could be done, this invariably leads to a busy waiting solution
where a process keeps obtaining locks and releasing them again until it
finally is able to get all of them.

The \emph{No Circular Waiting} approach
is to prevent a cycle from forming, with each
process waiting for the next process on the cycle.
We can do this by establishing an ordering among the
resources (in this case the forks) and, when needing more than one
resource, always acquiring them in order.  In the case of the philosopers,
they could prevent deadlock by always picking up the lower numbered fork
before the higher numbered fork, like so:

\vspace{1em}
\begin{code}
\begin{Verbatim}[xleftmargin=5mm,numbers=left]
    if left < right:
        lock(&forks[left]);
        lock(&(orks[right]);
    else:
        unlock(&forks[right]);
        unlock(&forks[left]);
    ;
\end{Verbatim}
\end{code}
\vspace{1em}

This completes all the Havender methods.
There is, however, another approach, which is sometimes called
\emph{deadlock avoidance}.
\index{deadlock avoidance}
In the case of the Dining Philosophers, we want to avoid the situation where each
diner picks up a fork.  So if we can prevent more than four diners from starting to
eat at the same time, then we can avoid the conditions for deadlock from ever
happening.
Figure~\ref{fig:dinersavoid} demonstrates this concept.  It uses a semaphore to
restrict the number of diners at any time to four.

@@

\section*{Exercises}
\begin{problems}
\item Figure~\ref{fig:bank} shows an implementation of a bank with various
accounts and transfers between those accounts.
Unfortunately, running the test reveals that it sometimes leaves unterminated
processes.  Can you fix the problem?
\end{problems}
